---
title: "Coursera Reproducible Research Assignment1"
output: html_document
---
Loading necessary libraries and read the data
```{r warning=F}
library(plyr)
library(lattice)
library(knitr)
activityData <- read.csv('activity.csv', header = TRUE, colClasses=c("numeric", "character", "numeric"))
activityData$date <- as.Date(activityData$date, format = "%Y-%m-%d")
activityData$interval <- as.factor(activityData$interval)
```
##### What is mean total number of steps taken per day?
Clean the data and calculate sums of steps perday
```{r}
steps_per_day <- aggregate(steps ~ date, activityData, sum)
colnames(steps_per_day) <- c("date","steps")
```

Histogram of total steps per day
```{r}
hist(steps_per_day$steps, breaks=15, main="Total number of steps taken each day", xlab="Total number of steps", col="red")
```


Average number of steps taken by the volunteers: `r mean(steps_per_day$steps)` and median:`r median(steps_per_day$steps)`

##### What is the daily activity pattern?
```{r}
intSteps <- aggregate(activityData$steps, by = list(interval = activityData$interval),FUN=mean, na.rm=TRUE)
intSteps$interval <- as.integer(levels(intSteps$interval)[intSteps$interval])
colnames(intSteps) <- c("interval", "steps")
```

Time series plot (i.e. type = 'l') of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r}
plot(intSteps$interval, intSteps$steps, type="l", main="Daliy activity pattern in 5 min interval",xlab="Interval", ylab="Average number of steps")
```

```{r}
maxInterval <- intSteps[intSteps$steps==max(intSteps$steps),]$interval
```

Maximum number of steps taken at `r maxInterval` minutes, i.e. at `r maxInterval %/% 60` hours  and `r maxInterval %% 60` minutes from the start of data collection time.

##### Imputing missing values.
Total numberof missing values in the given datare are `r sum(is.na(activityData))`

``` {r}
avgSteps <- aggregate(steps ~ interval, data = activityData, FUN = mean)
fillNA <- numeric()
for (i in 1:nrow(activityData)) {
  obs <- activityData[i, ]
  if (is.na(obs$steps)) {
    steps <- subset(avgSteps, interval == obs$interval)$steps
  } else {
    steps <- obs$steps
  }
  fillNA <- c(fillNA, steps)
}
newActivity <- activityData
newActivity$steps <- fillNA
tidyDataTotSteps <- aggregate(steps ~ date, data = newActivity, sum, na.rm = TRUE)
```

Plotting the histogram with imputed values
``` {r}
hist(tidyDataTotSteps$steps, breaks = 15, main = "Total steps by day", xlab = "Day", col = "red")
```

After filling the NA values with mean of the intervals, the average number of steps taken by the volunteers: `r round(mean(tidyDataTotSteps$steps),5)` and median:`r round(median(tidyDataTotSteps$steps), 5)`

##### Differences in activity patterns between weekdays and weekends
``` {r}
day <- weekdays(activityData$date)
dayType <- vector()
for (i in 1:nrow(activityData)) {
  if (day[i] == "Saturday"| day[i] =="Sunday") {
    dayType[i] <- "Weekend"
  } else {
    dayType[i] <- "Weekday"
  }
}
activityData$dayType <- dayType
activityData$dayType <- factor(activityData$dayType)
stepsDayType <- aggregate(steps ~ interval + dayType, data = activityData, mean)
names(stepsDayType) <- c("interval", "dayType", "steps")
```

Time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekdays or weekends (y-axis).
``` {r}
xyplot(steps ~ interval | dayType, stepsDayType, type = "l", layout = c(1, 2), 
         xlab = "Interval", ylab = "Number of steps")
```

